---
title: "P8108 Homework 3"
author: "Ryan Wei, rw2844"
date: "2022-09-28"
output:
  pdf_document:
    keep_tex:  true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r pkg, include=FALSE}
library(survival)
library(tidyverse)
library(biostat3)
library(ggfortify)
library(ggsurvfit)
library(gtsummary)
```

## Problem 1

```{r hand-draw-lifetable}
library(knitr)
library(kableExtra)
life_tab <- data.frame(Interval = c(1,2,3,4,5,6,7), 
                 Period = c("[0,4)","[4,8)","[8,12)","[12,16)","[16,20)","[20,24)","[24,$\\infty$)"),
                 d_i = c(2,1,0,1,1,1,0),
                 c_i = c(1,1,3,2,3,1,3),
                 n_i = c(20,17,15,12,9,5,3),
                 n_i_prime = c(19.5,16.5,13.5,11,7.5,4.5,1.5),
                 st = c(1.000,0.8974,0.8430,0.8430,0.7664,0.6642,0.5166),
                 ft = c(0.0257,0.0136,0,0.0192,0.0256,0.0369,NA),
                 ht = c(0.2703,0.0156,0,0.0238,0.0357,0.0625,NA),
                 se_st = c(0,0.06870,0.0833,0.0833,0.1053,0.1318,0.1657))
kable(life_tab, col.names = c("Interval", "Time Period", "$d_i$", "$c_i$", "$n_i$", "$n_i^{\\prime}$", "$S(t)$", "$f(t)$", "$h(t)$", "$se(S(t))$"), booktabs = T, escape = F, caption = "Life table estimate") %>% kable_styling(position = "center", latex_options = "hold_position") %>% footnote(general = "Survival functions estimated at the begining of each interval",footnote_as_chunk = T)
```
\newpage


## Problem 2

**Proof**

Since

$$
\begin{aligned}
\hat h(t_{mi}) & = \frac{d_i}{[(t_i -t_{i-1})(\frac{n_i^{\prime} - d_i}{2})]}\\
\hat f(t_{mi}) & = \frac{\hat S_L(t_{i-1})- \hat S_L(t_{i})}{t_i - t_{i-1}}\\
\hat S(t_{mi}) & = \frac{\hat S_L(t_{i-1}) + \hat S_L(t_{i})}{2}\\
\hat S_L(t_i) & = \hat S_L(t_{i-1})(1- \frac{d_i}{n_i^{\prime}})
\end{aligned}
$$
Therefore, based on the definition,
$$
\begin{aligned}
\hat h(t_{mi}) & = \frac{\hat f(t_{mi})}{\hat S(t_{mi})}\\
& = \frac{2\hat f(t_{mi})}{\hat S_L(t_{i-1}) + \hat S_L(t_{i})}\\
& = \frac{2(\hat S_L(t_{i-1}) - \hat S_L(t_{i}))}{(t_i - t_{i-1})(\hat S_L(t_{i-1}) + \hat S_L(t_{i}))}\\
& = \frac{2(\hat S_L(t_{i-1})\times\frac{d_i}{n_i^{\prime}})}{(t_i - t_{i-1})(\hat S_L(t_{i-1})\times\frac{2n_i^{\prime}-d_i}{n_i^{\prime}})}\\
& = \frac{d_i}{(n_i^{\prime}-\frac{d_i}{2})(t_i - t_{i-1})}
\end{aligned}
$$
\hfill $\square$

\newpage

## Problem 3

- Create life-table stratified by `rx`.

```{r life_table_rx1}
ovarian.rx1 <- ovarian[ovarian$rx == 1, ]
ovarian.rx2 <- ovarian[ovarian$rx == 2, ]
ovarian.lt1 <- lifetab2(Surv(futime, fustat)~1, data = ovarian.rx1, breaks =seq(0, 1500, 100))
ovarian.lt2 <- lifetab2(Surv(futime, fustat)~1, data = ovarian.rx2, breaks =seq(0, 1500, 100))
ovarian.lt1 %>% kable(booktabs = T, caption = "Life table, rx=1") %>% kable_styling(latex_options = c("HOLD_position"), font_size = 7)
```
```{r life_table_rx2}
ovarian.lt2 %>% kable(booktabs = T, caption = "Life table, rx=2") %>% kable_styling(latex_options = c("HOLD_position"), font_size = 7)
```


\newpage 

- Plot hazard function by `rx` based on life-table estimate

```{r life_table_hazard, out.width="80%"}
# hazard: the estimated hazard rate at the midpoint of the intervals.
ovarian.lt1$rx = as.factor(1)
ovarian.lt2$rx = as.factor(2)
ovarian.lt <- rbind(ovarian.lt1, ovarian.lt2)
ovarian.lt$tm <- (ovarian.lt$tstart + ovarian.lt$tstop)/2
ovarian.lt %>% 
  ggplot(data = ., aes(x = tm, y = hazard, group = rx, color = rx)) +
  geom_line() + geom_point()
```


- Plot K-M survival function by `rx`.

```{r km_surv_rx, out.width="80%"}
ovarian.kmfit <- 
  ovarian %>% 
  survfit(Surv(futime,fustat)~rx, data = .)

ovarian.kmfit %>% autoplot() +
  ylab(bquote(S(t))) +
  ggtitle("Kaplan-Meier survival curve \n by treatment")


#ovarian.kmfit %>% ggsurvfit()
```
\newpage

- What is the median survival time for each treatment group?

```{r median_surv}
km_median = 
  tibble(
    rx = as.factor(c(1,2)),
    mediansurvival = c("638 (268,--)", " -- (475,--)")
  )
km_median %>% kable(booktab = T, escape = F, col.names = c("rx", "Median Survival (95$\\%$ CI)"), caption = "Median survival time by K-M estimation") %>% kable_styling(latex_options = c("HOLD_position"))
```

- Compare survival function estimations between K-M and F-H methods.

```{r f_h_survival, out.width="60%"}
ovarian.fhfit <- 
  ovarian %>% 
  survfit(Surv(futime,fustat)~rx, data = ., type = "fh" )

par(mfrow=c(1,2))
ovarian.kmfit %>% 
  autoplot() +
  ylab(bquote(S(t))) +
  ggtitle("Kaplan-Meier survival curve \n by treatment")
ovarian.fhfit %>% 
  autoplot() + 
  ylab(bquote(S(t))) +
  ggtitle("Fleming-Harrington survival curve \n by treatment")
```


```{r survival_comp}
survival_tab <- tibble(
  time = ovarian.fhfit$time,
  kmsurv = ovarian.kmfit$surv,
  kmstd = ovarian.kmfit$std.err,
  fhsurv = ovarian.fhfit$surv,
  fhstd = ovarian.fhfit$std.err,
  rx = c(rep(1,13), rep(2,13))
)
survival_tab %>% filter(rx == 1) %>% dplyr::select(-rx) %>% kable(longtable = T, booktabs = T, caption = "Survival function estimations between K-M and F-H methods, rx = 1", col.names = c("Time", "Survival", "Survival Standard Error", "Survival", "Survival Standard Error"), digits = 4) %>% add_header_above(c(" ", "K-M estimation" = 2, "F-H estimation" = 2)) %>% kable_styling(latex_options = c("repeat_header"))

survival_tab %>% filter(rx == 2) %>% dplyr::select(-rx) %>% kable(longtable = T, booktabs = T, caption = "Survival function estimations between K-M and F-H methods, rx = 2", col.names = c("Time", "Survival", "Survival Standard Error", "Survival", "Survival Standard Error"), digits = 4) %>% add_header_above(c(" ", "K-M estimation" = 2, "F-H estimation" = 2)) %>% kable_styling(latex_options = c("repeat_header"))
```

- Describe your analyses and write conclusion based on you analyses.

1. In both stratums, there are 13 patients. However, there are more censored patients in group 2 (`rx = 2`) than group 1 (`rx = 1`). The hazard rate of group 1 is higher than group 2 for most of time.

2. According to different survival function estimations, both K-M curves and F-H curves show that the survival probability of group 2 is higher than group 1. 

3. The median survival time of group 1 is 638. However, group 2 doesn’t reach a median survival.

4. The estimation by K-M method and F-H method doesn’t differ too much.

## Reference

1. Edmunson, J.H., Fleming, T.R., Decker, D.G., Malkasian, G.D., Jefferies, J.A., Webb, M.J., and Kvols, L.K., Different Chemotherapeutic Sensitivities and Host Factors Affecting Prognosis in Advanced Ovarian Carcinoma vs. Minimal Residual Disease. Cancer Treatment Reports, 63:241-47, 1979.

## Appendix: Code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

